---
title: "HowToInvadeTheOccupiedNiche_Analyses"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Mathias Wegner, AWI"
date: "07/01/2022"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown of the analyses accompanying the manuscript 'Invading the occupied niche: aggregation behavior gives a generalist parasite the edge over an established specialist' by Marieke E. Feis, Leo Gottschalck, Lena Ruf, Franziska Theising, Felicitas Demann, K. Mathias Wegner. 
```{r, include = F}
library(vioplot)
library(nlme)
library(lmerTest)
library(lme4)
library(car)
library(multcomp)
```



Functions:
```{r functions}

#defining standard error
se = function(x)
{
	sd(x, na.rm=T)/sqrt(length(x))
}

#perform phase shift analysis
# parameters:
# x, y: numeric vectors for shifted correlation calucation
# left_shift, right_shift: integers for how many intervals series will be pushed left and right
# value:
# a list 1) R2: numric vector of all observed R2 values and 2) models: a list of the corresponding linear models
calcPhaseShift = function(x,y,left_shift,right_shift)
{
  
  models = list() #list of linear models for each time shift
  R2 = vector(mode = 'numeric') #vector of the correponding R2-values
  
  for(i in left_shift:right_shift)
  {
    if(i<0)
    {
      this.x = x[1:(length(x)-abs(i))]
      this.y = y[(abs(i)+1):length(y)]
    }
    if(i==0)
    {
      this.x = x
      this.y = y
    }
    
    
    if(i>0)
    {
      this.x = x[(abs(i)+1):length(x)]
      this.y = y[1:(length(y)-abs(i))]
    }
    
    mod = lm(this.x~this.y)
    modname = paste('mod',i,sep='_')
    models[[length(models)+1]] <- mod
    #print(summary(mod)) 
    R2 = c(R2,summary(mod)$r.squared)
    
    #rho = cor.test(x,y, method = 'spearman')
    #Rho = c(Rho,rho$estimate)
    
  }
  return(list(R2=R2,models=models))
}

```

##PART I: Field data
This part analyses the field data collected in Königshafen from 2013 until 2019.


```{r Figure1_data, include = F}
#Loading data and data manipulation
data = read.table('Mytilicola_export.txt', header = T)
data$M_infected_Mint = 0
data$M_infected_Mint[data$sum_int>0]=1

data$M_infected_Mori = 0
data$M_infected_Mori[data$sum_ori>0]=1

data$M_infected = 1
data$M_infected[data$M_infected_Mint == 0 & data$M_infected_Mori == 0] = 0

data$O_infected_Mori = 0
data$O_infected_Mori[data$O_Mori_sum>0]=1


data$propRepro_M_mint   = NA
data$propRepro_M_mint[complete.cases(data$sum_int) & data$sum_int>0] = 1- I(0.5^(data$sum_int[complete.cases(data$sum_int) & data$sum_int>0]-1))
data$propRepro_M_mori   = NA
data$propRepro_M_mori[complete.cases(data$sum_ori) & data$sum_ori>0] = 1- I(0.5^(data$sum_ori[complete.cases(data$sum_ori) & data$sum_ori>0]-1))
data$propRepro_O_mori   = NA
data$propRepro_O_mori[complete.cases(data$O_Mori_sum) & data$O_Mori_sum>0] = 1- I(0.5^(data$O_Mori_sum[complete.cases(data$O_Mori_sum) & data$O_Mori_sum>0]-1))
```


```{r Figure1_prevalences, include = F}
#Calculate mean prevalences and their associated errors
xmon=c(0,12,24,36,48,60,72,84) #84 = 2020
M_inf = aggregate(M_infected~monthnum+Year, data = data, mean)
M_inf$x_month =M_inf$monthnum+xmon[as.numeric(as.factor(M_inf$Year))]

M_Mint_prev = aggregate(M_infected_Mint~monthnum+Year, data = data, mean)
M_Mint_prev$x_month =M_Mint_prev$monthnum+xmon[as.numeric(as.factor(M_Mint_prev$Year))]

M_Mori_prev = aggregate(M_infected_Mori~monthnum+Year, data = data, mean)
M_Mori_prev$x_month =M_Mori_prev$monthnum+xmon[as.numeric(as.factor(M_Mint_prev$Year))]

O_Mori_prev = aggregate(O_infected_Mori~monthnum+Year, data = data, mean)
O_Mori_prev$x_month =O_Mori_prev$monthnum+xmon[as.numeric(as.factor(M_Mint_prev$Year))]

M_Mint_prev$M_infected_Mori = M_Mori_prev$M_infected_Mori
```


Phase shift calculations for Figure 1C,D
```{r Figure1_phase shift calculations, echo = T}
#reducing data set to times of co-occurence
after2015 = subset(M_Mint_prev, M_Mint_prev$Year>2015)

#phase shift
x_phase = -3:4
MintMori_shift = calcPhaseShift(after2015$M_infected_Mint,after2015$M_infected_Mori,-3,4)
bestMod_Index = which(MintMori_shift$R2==max(MintMori_shift$R2))
bestMod = MintMori_shift$models[[bestMod_Index]]
print(paste('Best correlation at shift:',x_phase[bestMod_Index]))
summary(bestMod)

```

Plot Figure 1 B,C,D

```{r Plot 1B,C,D, echo = F}
layout(matrix(c(1,1,1,1,2,3),2,3))
plot(M_Mint_prev$M_infected_Mint~M_Mint_prev$x_month,ylab = 'Prevalence', xaxt = 'n', xlab = 'Year', pch = 16, type = 'b', col = 'blue', lwd = 3, ylim = c(0,1), main = 'Prevalence', cex = 1.5)
axis(1,at = xmon +1, c('2013','2014','2015','2016','2017','2018','2019','2020'), cex = 1.5)
points(M_Mori_prev$M_infected_Mori~M_Mori_prev$x_month, pch = 16, type = 'b', col = 'red', lwd = 3, cex = 1.5)
points(O_Mori_prev$O_infected_Mori~O_Mori_prev$x_month, pch = 1, type = 'b', col = 'dark red', lty = 2, lwd = 3, cex = 1.5)

points(M_inf$M_infected ~M_inf$x_month, type = 'l', lty = 2,lwd = 2)

plot(x_phase,MintMori_shift$R2, type = 'l', lwd = 3, main = 'Phase shift (2016-2019)', xlab = 'Phase shift M. intestinalis')
abline(v=0, lty = 3)

plot(bestMod$model$this.x ~ bestMod$model$this.y,  pch = 16, ylab = 'Prev. M. intestinalis', xlab = 'Prev. M. orientalis', main = 'Correlation +1', cex = 1.5)
abline(bestMod, lwd = 2)

```

Calculate dispersion and reproductive opportunities
'M_Mint_abundance' is a sampling interval aggregate of M. intestinalis (Mint) abundances in mussels (M), Mori is M. orientalis, and O is oyster. 
Different stats are calculated for each host parasite combination (M_Mint, M_mori, O_Mori)
For all 3 combinations data is merged into M-Mint_abundance

```{r calculate Dispersion, include =F}
#Dispersion 
M_Mint_abundance=aggregate(sum_int~monthnum+Year, data = data,mean,na.rm=T)
M_Mint_var = aggregate(sum_int~monthnum+Year, data = data,var,na.rm=T)
M_Mint_abundance$var= M_Mint_var$sum_int
M_Mint_abundance$log_ab = log(M_Mint_abundance$sum_int)
M_Mint_abundance$log_var = log(M_Mint_var$sum_int)
M_Mint_abundance$log_var_over_log_ab = M_Mint_abundance$log_var/M_Mint_abundance$log_ab
M_Mint_abundance$var_over_mean = M_Mint_abundance$var/M_Mint_abundance$sum_int
M_Mint_abundance$x_month=M_Mint_prev$x_month




M_Mori_abundance =aggregate(sum_ori~monthnum+Year, data = data,mean,na.rm=T)
M_Mori_var = aggregate(sum_ori~monthnum+Year, data = data,var,na.rm=T)
M_Mori_abundance$var= M_Mori_var$sum_ori
M_Mori_var_over_mean = M_Mori_abundance$var/M_Mori_abundance$sum_ori
M_Mori_var_over_mean=c(rep(NA,25),M_Mori_var_over_mean[4:14])
M_Mint_abundance$M_Mori_var_over_mean=M_Mori_var_over_mean #merging all data into M-Mint_abundance



O_Mori_abundance =aggregate(O_Mori_sum~monthnum+Year, data = data,mean,na.rm=T)
O_Mori_var =aggregate(O_Mori_sum~monthnum+Year, data = data,var,na.rm=T)
O_Mori_abundance$var=O_Mori_var$O_Mori_sum
O_Mori_abundance$var_over_mean = O_Mori_abundance$var/O_Mori_abundance$O_Mori_sum
O_Mori_var_over_mean = c(rep(NA,26),O_Mori_abundance$var_over_mean[2:11])
M_Mint_abundance$O_Mori_var_over_mean = O_Mori_var_over_mean


#ReproSuccess
M_Mint_Repro = aggregate(propRepro_M_mint~monthnum+Year, data = data,mean,na.rm=T)
M_Mori_Repro = aggregate(propRepro_M_mori~monthnum+Year, data = data,mean,na.rm=T)
O_Mori_Repro = aggregate(propRepro_O_mori~monthnum+Year, data = data,mean,na.rm=T)

M_Mint_abundance$repro_M_Mint = M_Mint_Repro$propRepro_M_mint #merging all data into M-Mint_abundance
M_Mint_abundance$repro_M_Mori = c(rep(NA,26),M_Mori_Repro$propRepro_M_mori)
M_Mint_abundance$repro_O_Mori = c(rep(NA,26),O_Mori_Repro$propRepro_O_mori)

means = c(M_Mint_abundance$sum_int,M_Mori_abundance$sum_ori,O_Mori_abundance$O_Mori_sum)
vars = c(M_Mint_var$sum_int,M_Mori_var$sum_ori,O_Mori_var$O_Mori_sum)
repro = c(M_Mint_Repro$propRepro_M_mint, NA,NA,NA,M_Mori_Repro$propRepro_M_mori,NA,O_Mori_Repro$propRepro_O_mori)
hostPar = c(rep('M_Mint',length(M_Mint_abundance[[1]])),rep('M_Mori',length(M_Mori_abundance[[1]])),rep('O_Mori',length(O_Mori_abundance[[1]])))

#data frame for dispersion parameters
disp =data.frame(means=means,vars=vars,hostPar=hostPar, repro = repro)
disp=subset(disp,disp$vars>0)
disp$ln_mean =log(disp$means)
disp$ln_var = log(disp$vars)
disp$varToMean = disp$vars/disp$means

```


Analysis for Figure 2B: Aggregation in different parasite-host combinations

```{r model Fig2b, echo = T}
#analysis
#test for species differences accounting for different variances
fit.gls <- gls(varToMean~hostPar, weights = varIdent(form = ~ 1 | hostPar), data = disp)
summary(fit.gls)
anova(fit.gls)
```


Analysis Figure 2C: Chances of encountering a mate

```{r model Fig2C, echo = T}
#### prop Repro 2C
#regression before M. orientalis arrival in 2016
M_Mint_abundance_before = subset(M_Mint_abundance,M_Mint_abundance$x_month < 35)
M_Mint_abundance_after = subset(M_Mint_abundance,M_Mint_abundance$x_month > 35)

M_mint_repro_mod_before = lm(M_Mint_abundance_before$repro_M_Mint~ M_Mint_abundance_before$x_month)
print(summary(M_mint_repro_mod_before))

M_mint_repro_mod_after = lm(M_Mint_abundance_after$repro_M_Mint~ M_Mint_abundance_after$x_month)
print(summary(M_mint_repro_mod_after))


M_mori_repro_mod = lm(repro_M_Mori~x_month, data = M_Mint_abundance)
print(summary(M_mori_repro_mod))


O_mori_repro_mod = lm(repro_O_Mori~x_month, data = M_Mint_abundance)
print(summary(O_mori_repro_mod))

```

Plotting of Figure 2


```{r plot Figure2, echo = F}
#Intensity distributions 2A
layout(matrix(c(1,2,3,4,4,4,4,4,4,5,5,5,5,5,5),3,5))
par(mar=c(5,4,3,2))

hist(data$sum_int[complete.cases(data$sum_int) & data$sum_int>0], col = 'blue',breaks = 0:46, xlab = '', ylab = '',main = 'M. intestinalis in mussel')
abline(v=mean(data$sum_int[complete.cases(data$sum_int) & data$sum_int>0]), col = 'blue', lty = 2)
text(25,100,paste('n = ',length(data$sum_int[complete.cases(data$sum_int) & data$sum_int>0])))

hist(data$sum_ori[complete.cases(data$sum_ori) & data$sum_ori >0], col = 'red',breaks = 0:46,xlab = '', ylab = '', main = 'M. orientalis in mussel')
abline(v=mean(data$sum_ori[complete.cases(data$sum_ori) & data$sum_ori >0]), col = 'red', lty = 2)
text(25,30,paste('n = ',length(data$sum_ori[complete.cases(data$sum_ori) & data$sum_ori >0])))


hist(data$O_Mori_sum[complete.cases(data$O_Mori_sum) & data$O_Mori_sum>0], col = 'darkred',breaks = 0:46,xlab = '', ylab = '', main = 'M. orientalis in oyster')
abline(v=mean(data$O_Mori_sum[complete.cases(data$O_Mori_sum) & data$O_Mori_sum>0]), col = 'darkred', lty = 2)
text(25,10,paste('n = ',length(data$O_Mori_sum[complete.cases(data$O_Mori_sum) & data$O_Mori_sum>0])))

# Variance to mean plot 2B
vioplot(disp$varToMean~disp$hostPar, col = c('blue','red','darkred'), ylab = 'Variance to mean ratio (VMR)', main = 'Aggregation of parasites', xlab='Host - Parasite Combination', cex = 1.3)
abline(h=1, lty = 2, lwd = 2)

#plot repro success 2C
plot(M_Mint_abundance$x_month,M_Mint_abundance$repro_M_Mint, xaxt = 'n', ylab = 'Likelihood of finding a mate', xlab = 'Year', type = 'p', pch = 16, col = 'blue', main = 'Mate finding')
axis(1,at = xmon [2:8], c('2013','2014','2015','2016','2017','2018','2019'))

#adding regression lines
points(M_Mint_abundance_before$x_month[c(1,25)],predict(M_mint_repro_mod_before)[c(1,25)], type = 'l',lwd = 2, col = 'blue', lty = 2)
points(M_Mint_abundance_after$x_month[c(1,11)],predict(M_mint_repro_mod_after)[c(1,11)], type = 'l',lwd = 2, col = 'blue', lty = 2)

#Mori in mussel
points(repro_M_Mori~x_month, data = M_Mint_abundance, col = 'red', pch = 16)
abline(M_mori_repro_mod, col = 'red', lwd = 2, lty = 2)

#Mori in oyster
points(repro_O_Mori~x_month, data = M_Mint_abundance, col = 'dark red', pch = 1, cex =1.3, lwd = 1.5)
abline(O_mori_repro_mod, col = 'darkred', lwd = 2, lty = 2)
```


##Part II: Simultaneous and Sequential infections (Fig 3A)
This part analyses the simultaneous and sequential infection data, corresponding to Figure 3
Simultaneous infections (Fig 3A)
```{r simultaneous infections read data, include = F }

#### Import dataset ####
dataSim = read.table('simultaneousInfection.txt', header = T)
dataSim$infRate2 = dataSim$Parasite_count/dataSim$exposureDose

```



```{r simultaneous infections analyses}
#############################################################################################################
# test by binomial GLMM
#############################################################################################################

mod = glmer(cbind(Parasite_count,(exposureDose-Parasite_count))~Parasite * treatment + (1|Mussel_name), data = dataSim, family = 'binomial',control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
print(summary(mod))

#############################################################################################################
# pairwise comparisons
#############################################################################################################
dataSim$PxT = interaction(dataSim$Parasite,dataSim$treatment)
compMod = glmer(cbind(Parasite_count,(exposureDose-Parasite_count))~PxT + (1|Mussel_name), data = dataSim, family = 'binomial',control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
print(summary(compMod))
print(summary(glht(compMod, linfct = mcp(PxT = 'Tukey'))))

meanInf = aggregate( infRate2 ~ treatment*Parasite, data = dataSim, mean)
meanSE = aggregate( infRate2 ~ treatment*Parasite, data = dataSim, se)
```



##Sequential Infections (Fig 3B)

```{r sequenttial infections read data, include = F}


#############################################################################################################
#############################################################################################################
#
#                               Sequential Infections
#
#############################################################################################################
#############################################################################################################


#read data
dataSeq = read.table('SequentialInfection.txt', header =T)
dataSeq$Mussel = factor(dataSeq$Mussel)
dataSeq$HomHet = factor(dataSeq$HomHet, levels = c('single','homolog','heterolog'))

#calculate Infection rate
dataSeq$infRate = dataSeq$inf/dataSeq$dose


dataSeq$Inf_combo = factor(dataSeq$Inf_combo, levels = c("Mintcontrol","controlMint","MintMint","MintMori","MoriMint","MoriMori","Moricontrol","controlMori"))
dataSeq$infection = 'single'
dataSeq$infection[dataSeq$HomHet == 'homolog' | dataSeq$HomHet =='heterolog'] = 'coinfection'
```

```{r sequenttial infections analysis}
#############################################################################################################
# test by binomial GLMM
#############################################################################################################

modSeq = glmer(cbind(inf,(dose-inf))~ Parasite * round * HomHet + (1|Mussel), data = dataSeq, family = 'binomial',control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
#summary(modSeq)
Anova(modSeq, type = 'III')
```

### Pairwise comparisons of Sequential infections

```{r sequential infections - pairwise, echo = F}
#############################################################################################################
# pairwise comparisons
#############################################################################################################
dataSeq$PxRxHH = interaction(dataSeq$Parasite, dataSeq$round, dataSeq$HomHet)
dataSeq$PxRxHH = as.character(dataSeq$PxRxHH)
dataSeq$PxRxHH[dataSeq$PxRxHH == 'Mint.inf1.homolog'] = 'Mint.homolog'
dataSeq$PxRxHH[dataSeq$PxRxHH == 'Mint.inf2.homolog'] = 'Mint.homolog'
dataSeq$PxRxHH[dataSeq$PxRxHH == 'Mori.inf1.homolog'] = 'Mori.homolog'
dataSeq$PxRxHH[dataSeq$PxRxHH == 'Mori.inf2.homolog'] = 'Mori.homolog'
dataSeq$PxRxHH = factor(dataSeq$PxRxHH)

compModSeq = glmer(cbind(inf,(dose-inf))~PxRxHH + (1|Mussel), data = dataSeq, family = 'binomial',control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
print(summary(compModSeq))
print(summary(glht(compModSeq, linfct = mcp(PxRxHH = 'Tukey'))))

```

Plotting Figure 3

```{r Figure3, echo = F}
layout(matrix(c(1,2,2),1,3))
par(mar=c(5,6,3,0))
vioplot(infRate2 ~ treatment, data=dataSim[dataSim$Parasite == 'Mint',], col = 'blue', ylim = c(0,1), xlim = c(0.5,4), xaxt='n', xlab = '', ylab = '', main = 'Simultaneous Infection', cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5, cex = 1.5)
vioplot(infRate2 ~ treatment, data=dataSim[dataSim$Parasite == 'Mori',], add = T, col = 'red', at = c(2.5,3.5), xaxt = 'n', cex = 1.5)
axis(1,at = c(1,2.25,3.5), labels = c("","",""))
mtext(c('Mint','Mint-Mori', 'Mori'), side = 1,line = 1, at = c(1,2.25,3.5))
mtext("Infection rate", side = 2, line = 3, cex = 1.5)

par(mar=c(5,0,3,1))
x=vioplot(infRate ~ factor(Inf_combo,levels = c("Mintcontrol","controlMint","MintMint","MintMori","MoriMint")), data=dataSeq[dataSeq$Parasite == 'Mint',], col = 'blue', ylim = c(0,1), at=c(1,2,3.5,4.5,6.5), xlim = c(0.5,10.5), xaxt = 'n', main = 'Sequential Infections', ylab="", xlab ="", cex = 1.5, cex.main = 1.5, yaxt ='n')
y=vioplot(infRate ~ factor(Inf_combo,levels = c("MintMori","MoriMint","MoriMori","Moricontrol","controlMori")), data=dataSeq[dataSeq$Parasite == 'Mori',], col = 'red', ylim = c(0,1), add=T,at=c(5,6,7.5,9,10), xaxt = 'n', cex = 1.5)
axis(1,at = c(1,2,3.5,4.75,6.25,7.5,9,10), labels = c('','','','','','','',''))
mtext(c('Mint','-','Mint','Mint','Mori','Mori','Mori','-'),side = 1, line = 1, at = c(1,2,3.5,4.75,6.25,7.5,9,10),cex = 1)
mtext(c('-','Mint','Mint','Mori','Mint','Mori','-','Mori'), side = 1, line = 2, at = c(1,2,3.5,4.75,6.25,7.5,9,10), cex = 1)


```

##PART III:  Host dependence of infectivity

Tests whether the source species (mussel or oyster) has an effect on the infectitivity of parasite offspring in both species

```{r host source dependence read data, include =F}
dataHost = read.table('hostSource.txt',header =T, stringsAsFactors = T)
dataHost$infCombo = paste(dataHost$Treatment,dataHost$HostSpecies,sep = '')
dataHost$infCombo = factor(dataHost$infCombo, levels = c('MiM_edulis','MiC_gigas','MoMeM_edulis','MoMeC_gigas','MoCgM_edulis','MoCgC_gigas'))

dataHostNoMiCG = subset(dataHost,dataHost$infCombo != 'MiC_gigas')
```

```{r host source dependence analysis}
parental.mod = glm(cbind(M_Total,(Cop_infiltrated-M_Total))~Treatment, data = dataHostNoMiCG, family = 'binomial')
summary(parental.mod)
anova(parental.mod,test = 'LRT')

dataHost_Mori = subset(dataHostNoMiCG,dataHostNoMiCG$Treatment != 'Mi')
hostMori.mod = glm(cbind(M_Total,(Cop_infiltrated-M_Total))~Treatment * HostSpecies, data = dataHost_Mori, family = 'binomial')
summary(hostMori.mod)


```

Plotting Figure 4

```{r Figure 4, echo = F}
vioplot(inf_rate ~ infCombo, data = dataHost, col = c('blue','blue','red','dark red','red','dark red'), at = c(1,2,3.5,4.5,6,7), xaxt = 'n', xlab ='', ylab = 'Infection rate')
axis(1, at =c(1,2,3.5,4.5,6,7), labels= rep('',6))
mtext(c('M.int','M.int','M.ori','M.ori','M.ori','M.ori'),side = 1, line = 1,  at = c(1,2,3.5,4.5,6,7),cex = 1,col = c('blue','blue','red','dark red','red','dark red'))
mtext('Parasite', side = 1, line = 1, at = -0.0)

mtext(c('mussel','mussel','mussel','oyster','mussel','oyster'),side = 1, line = 2,  at = c(1,2,3.5,4.5,6,7),cex = 1,col = c('blue','blue','blue','red','blue','red'))
mtext('Infection host', side = 1, line = 2, at = -0.0)

mtext(c('mussel','oyster','mussel','mussel','oyster','oyster'),side = 1, line = 3,  at = c(1,2,3.5,4.5,6,7),cex = 1,col = c('blue','red','blue','blue','red','red'))
mtext('Source host', side = 1, line = 3, at = 0)

```

## PART IV: Infection preference

tests whether the prior presence of other con- or heterospecific infections changes the infection choice when offered an infected vs. an uninfected host

```{r Infection Choice read data, echo = F}
dataChoice = read.table('InfectionChoice.txt', header =T, stringsAsFactors = T)

#create data set for only mytilus
mytilus = subset(dataChoice,dataChoice$Host=='Mytilus')
mytilus = subset(mytilus, mytilus$dead_inf =="alive" & mytilus$dead_naive == "alive")
mytilus$checkSum = mytilus$SecondIntensity_inf + mytilus$SecondIntensity_naive #max = 23 passt

#calculate proportions of infections
mytilus$FirstInfection_prop = mytilus$FirstIntensity/mytilus$FirstIngested
mytilus$SecondInfection_prop = (mytilus$SecondIntensity_inf + mytilus$SecondIntensity_naive)/mytilus$SecondIngested
mytilus$SecondInfection_inf_prop = mytilus$SecondIntensity_inf/mytilus$SecondIngested
mytilus$SecondInfection_naive_prop = mytilus$SecondIntensity_naive/mytilus$SecondIngested


#calculate difference between proportions inf-naive
mytilus$SecondInfection_prop_difference = mytilus$SecondInfection_inf_prop - mytilus$SecondInfection_naive_prop

mytilus$SecondInfection_difference = mytilus$SecondIntensity_inf - mytilus$SecondIntensity_naive

mytilus$SecondInfection_prop_sum = mytilus$SecondInfection_inf_prop + mytilus$SecondInfection_naive_prop

#data set without previous infections
mytilus_noPreinf = subset(mytilus,mytilus$pre_infected_inf == 'none')
```

```{r Infection Choice analysis}
# linear model with infection choice parasite species and homologous vs heterologous infection as factor
mod_hom_het = lm(SecondInfection_prop_difference~ FirstIntensity + SecondInfection * hom_het  , data = mytilus_noPreinf)
#anova table
anova(mod_hom_het)
summary(mod_hom_het)

```

Plot Figure 5

```{r Figure 5, echo = F}
#calculate mean and standard error for plotting
mean_dif = aggregate(SecondInfection_prop_difference~FirstInfection * SecondInfection, data = mytilus_noPreinf, mean)
se_dif = aggregate(SecondInfection_prop_difference~FirstInfection * SecondInfection, data = mytilus_noPreinf, se)

mean_sum = aggregate(SecondInfection_prop_sum~FirstInfection * SecondInfection, data = mytilus_noPreinf, mean)
se_sum = aggregate(SecondInfection_prop_sum~FirstInfection * SecondInfection, data = mytilus_noPreinf, se)

#make a bar plot, x = gives the middle of each bar on the x-axis
par(mar=c(5,5,3,2)) #increases the margins
x=barplot(mean_dif$SecondInfection_prop_difference, ylim = c(-0.3,0.5), col = c( 'blue', 'blue','red','red'), density = c(NA,30,30,NA))
abline(h=0) #draws horizontal line at 0
#add error bars
arrows(x,mean_dif$SecondInfection_prop_difference - se_dif$SecondInfection_prop_difference, x , mean_dif$SecondInfection_prop_difference + se_dif$SecondInfection_prop_difference, length = 0)

meanFirstRates = aggregate(FirstInfection_prop ~ FirstInfection + SecondInfection, data = mytilus_noPreinf, mean)
meanSecondRates = aggregate(SecondInfection_prop ~ FirstInfection + SecondInfection, data = mytilus_noPreinf, mean)


#add axis labels
mtext('Primary infection',side =1, line = 1, at = 0, cex = 1)
mtext('Preference infection',side =1, line = 2, at = 0, cex = 1)
mtext(c(paste('M. int (',round(meanFirstRates$FirstInfection_prop[1],2),')', sep = ""),
        paste('M. ori (',round(meanFirstRates$FirstInfection_prop[2],2),')', sep = ""),
        paste('M. int (',round(meanFirstRates$FirstInfection_prop[3],2),')', sep = ""),
        paste('M. ori (',round(meanFirstRates$FirstInfection_prop[4],2),')', sep = "")),
        side =1, line = 1, at = x, col = c('blue','red','blue','red'))
mtext(c(paste('M. int (',round(meanSecondRates$SecondInfection_prop[1],2),')', sep = ""),
        paste('M. int (',round(meanSecondRates$SecondInfection_prop[2],2),')', sep = ""),
        paste('M. ori (',round(meanSecondRates$SecondInfection_prop[3],2),')', sep = ""),
        paste('M. int (',round(meanSecondRates$SecondInfection_prop[4],2),')', sep = "")),
        side =1, line = 2, at = x, col = c('blue','blue','red','red'))
mtext('infected', side = 2, line = 2, at = 0.25)
mtext('not infected', side = 2, line = 2, at = -0.15)
mtext('Preference', side = 2, line = 3, at = 0, cex =1.2)

#add arrows to margin 
arrows(-0.3,0,-0.3,0.2, xpd = T)
arrows(-0.3,0,-0.3,-0.1, xpd = T)
arrows(-0.35,0,-0.25,0,length = 0, xpd =T)



```




